---
title: "FEMA Data Doc"
output: html_notebook
---

## Question: Is there an emerging pattern of increasing delays for federal disaster response?

## Data Sets: Disaster Declarations Summaries v2: <https://drive.google.com/file/d/1YhwtLE-e428HYR-CLruthPc3VHoWb2cg/view?usp=sharing>

## Methodology

FEMA's Disaster Declarations Summaries v2, a publicly available dataset spanning the Eisenhower administration to the present, was used for this analysis, where each row represents a county-level disaster designation. The variable days_to_declaration measures the number of calendar days between incidentBeginDate and declarationDate using difftime() in R, restricted to Major Disaster Declarations only (declarationType == "DR") as these usually indicate more broad federal assistance.

Because each row is a county designation rather than a unique disaster, the data needed to be cleaned. A check of ~3,000 state-incident combinations found only 13 with variation in declaration dates within a single state, which were nearly all attributable to tribal nations receiving separate, typically later, declarations. This analysis groups at the state level and takes the earliest declaration date within each state-incident group — tribal declarations tend to come later and are treated as a separate phenomenon. County-level rows are kept to report minimum, median, and maximum lag per state rather than reducing to a single number. Median is used as the primary statistic over mean to avoid a single outlier county disproportionately skewing state-wide data.

Hurricane Helene is isolated by incidentId rather than name string, as a string search for "helene" returns rows from an older unrelated storm since FEMA reuses storm names.

For the overall trend, all DR declarations were filtered to hurricanes and tropical storms only, then days_to_declaration was plotted against incidentBeginDate with a LOESS smoothed trend line. Response times were also broken down by decade, reporting both mean and median to show how the distribution has shifted over time.

For tribal declarations, the data was filtered using both the tribalRequest flag and a string search across designatedArea for keywords like "indian nation," "indian reservation," "pueblo," and "tribe of." Tribal and non-tribal declaration lag were then compared within the same state-incident to identify delays — Florida is highlighted specifically because it provides a direct within-state comparison across multiple storms.

Two key limitations throughout are that the lag clock starts at incident begin, but FEMA cannot declare until a state governor formally requests it, meaning some delay may reflect state-level decisions rather than federal response time. Additionally, data manually entered before the 1998 launch of the National Emergency Management Information System may be subject to a small percentage of human error.
---

```{r setup}
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(tidyr)
```

```{r load}
df_FEMA_OG <- read.csv('FEMA_disaster_declarations.csv')
#each row represents one county-level declaration -- no modifications to the data were made before loading into R
```

```{r clean-dates}
# converting all three date columns from character strings to proper Date objects so
# difftime() and any date math works correctly below
df_FEMA_clean <- df_FEMA_OG %>%
  mutate(
    declarationDate   = as_date(declarationDate),
    incidentBeginDate = as_date(incidentBeginDate),
    incidentEndDate   = as_date(incidentEndDate)
  )
```

---

## Checking incidentId vs. string filtering (Professor's note)

Filtering by name string catches an older storm also named Helene. Used `incidentId`.

```{r helene-id-vs-string}
helene_by_id <- df_FEMA_OG %>%
  filter(incidentId == '2024092301')

helene_by_string <- df_FEMA_OG %>%
  filter(grepl("helene", declarationTitle, ignore.case = TRUE))

#since helene by string has more, let's see what it has that's not in the list filtered by ID number

helene_missing_in_id <- helene_by_string %>%
  filter(!id %in% helene_by_id$id)

## and vice versa just in case 

helene_missing_in_string <- helene_by_id %>%
  filter(!id %in% helene_by_string$id)


#filtering using the string includes rows from another storm named helene before!!

cat("Rows caught by string but not ID (older storm):", nrow(helene_missing_in_id), "\n")
cat("Rows caught by ID but not string:", nrow(helene_missing_in_string), "\n")
```

---

## Days to Declaration — All DR Declarations

```{r all-dr}
# restricting to Major Disaster Declarations and creating column that calculates days between incident start and declaration
all_DR_data <- df_FEMA_clean %>%
  filter(declarationType == 'DR') %>%
  mutate(days_to_declaration = as.numeric(difftime(declarationDate, incidentBeginDate, units = "days")))
```

```{r dr-states-incidents}
# check whether declaration dates vary within a state for a given incident
# If count == 1 for all rows, every county in that state-incident got the same date
# only 13 out of ~3,000 state-incident combos have variation

DR_states_incidents <- all_DR_data %>%
  select(femaDeclarationString, state, incidentId, incidentBeginDate, declarationDate) %>%
  unique() %>%
  group_by(state, incidentId) %>%
  summarise(count = n(), .groups = "drop")

# the 13 exceptions
DR_states_incidents %>%
  filter(count > 1) %>%
  arrange(desc(count))
```

```{r check-exceptions-tribal}
# nearly all exceptions appear to be tribal declarations with a
# later date.

exception_ids <- DR_states_incidents %>%
  filter(count > 1) %>%
  pull(incidentId)

# pulling the full rows for each exception to look at designatedArea and tribalRequest
# and confirm that late dates are tribal rather than errors
all_DR_data %>%
  filter(incidentId %in% exception_ids) %>%
  select(femaDeclarationString, state, incidentId, designatedArea,
         incidentBeginDate, declarationDate, days_to_declaration, tribalRequest) %>%
  unique() %>%
  arrange(incidentId, declarationDate)
```

Since within-state date variation is minimal and largely driven by tribal declarations,
I grouped at the state level for the main analysis.

```{r dr-state-level}
# one row per state-incident: take the earliest declaration date within each group
# (tribal declarations tend to be later and are a separate phenomenon)

DR_state_level <- all_DR_data %>%
  select(femaDeclarationString, state, incidentId, incidentType,
         declarationTitle, incidentBeginDate, declarationDate,
         days_to_declaration, tribalRequest, fyDeclared) %>%
  unique() %>%
  group_by(state, incidentId) %>%
  slice_min(declarationDate, n = 1, with_ties = FALSE) %>%  # with_ties = FALSE avoids duplicate rows when two counties share the exact same earliest date
  ungroup()
```

---

## Hurricane Helene

```{r helene-dr}
# isolating Helene by incidentId rather than name string (see helene-id-vs-string chunk above)
helene_data_DR <- df_FEMA_clean %>%
  filter(declarationType == 'DR') %>%
  filter(incidentId == '2024092301') %>%
  mutate(days_to_declaration = as.numeric(difftime(declarationDate, incidentBeginDate, units = "days")))
```

```{r helene-state-summary}
# summarizing at the state level rather than county to get a cleaner picture —
# min/max/median gives a sense of spread within each state without flattening it to one number
helene_state_summary <- helene_data_DR %>%
  group_by(state) %>%
  summarise(
    n_counties = n(),
    min_lag    = min(days_to_declaration),
    median_lag = median(days_to_declaration),
    max_lag    = max(days_to_declaration),
    .groups    = "drop"
  ) %>%
  arrange(median_lag)

helene_state_summary

write.csv(helene_state_summary, "helene_state_summary.csv")

```

```{r helene-plot}
# I heavily used the help of Claude here to help me chart what I found visually. I think for KY and WV being a lot later that might be because of delayed effects? Or maybe that's something to look further into.
# bar height shows median lag in response. States where min and max are far apart
# probably need closer inspection where one county may have had an unusually delayed declaration
helene_state_summary %>%  
ggplot(aes(x = reorder(state, -median_lag), y = median_lag, fill = median_lag)) +
  geom_col() +
  geom_text(aes(label = median_lag), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +  # color encodes severity of delay — longer lags shade toward red
  labs(
    title    = "Hurricane Helene: Median Days to DR Declaration by State",
    subtitle = "Incident began 2024-09-23; bar shows median, see helene_state_summary for min/max",
    x        = NULL,
    y        = "Median Days to Declaration"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
##Tribal Declarations

```{r comparable-storm-names}
comparable_names <- c("camille", "floyd", "frances", "ivan", "jeanne",
                      "harvey", "ida", "matthew","milton", "helene")
#removed this earlier since analysis on comparable storms did not make it into story. pasting it back here to not throw a wrench in code below
```

```{r tribal}
# double-filtering for tribal designations: first by tribalRequest,
# then by string-matching designatedArea for common tribal naming conventions
tribal_DR <- all_DR_data %>%
  filter(
  tribalRequest == 1 | # tribalRequest is a binary flag; 1 = tribal declaration
  str_detect(str_to_lower(designatedArea), "indian nation|indian reservation|indian tribe|band of cherokee|band of indian|pueblo|tribe of")
)%>%
  filter(
    str_detect(str_to_lower(declarationTitle),
               paste(comparable_names, collapse = "|")) | incidentId == "2024092301"  # incidentId added as fallback since "helene" is in comparable_names but this makes the Helene filter more specific
  ) %>%
  group_by(state, designatedArea, declarationTitle, fyDeclared) %>%
  summarise(
    n_declarations = n(),
    min_lag        = min(days_to_declaration),
    median_lag     = median(days_to_declaration),
    max_lag        = max(days_to_declaration),
    .groups        = "drop"
  ) %>%
  arrange(desc(median_lag))

tribal_DR
```


```{r tribal-plot}
tribal_DR %>%
  mutate(area_label = paste0(designatedArea, " (", declarationTitle, ")")) %>%  # combining area and storm name into one label so each bar is fully identifiable without a legend
  ggplot(aes(x = reorder(area_label, median_lag), y = median_lag, fill = median_lag)) +
  geom_col() +
  geom_text(aes(label = median_lag), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Declaration Lag for Tribal Areas",
    subtitle = "Helene + comparable storms",
    x        = NULL,
    y        = "Median Days to Declaration"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 7))
```

```{r more tribal}
# adding in tribal to the group_by
DR_state_tribal_level <- all_DR_data %>%
  select(femaDeclarationString, state, incidentId, incidentType,
         declarationTitle, incidentBeginDate, declarationDate,
         days_to_declaration, tribalRequest, fyDeclared) %>%
  mutate(tribalRequest = ifelse(tribalRequest == 1, 'Days - Tribal', 'Days - Not Tribal')) %>%  # converting binary flag to readable labels for use as pivot column names below
  unique() %>%
  group_by(state, incidentId, tribalRequest) %>%
  slice_min(declarationDate, n = 1, with_ties = FALSE) %>%  # taking earliest date within each state-incident-tribalRequest group, same logic as DR_state_level
  ungroup()

# finding when more than one declaration for a given state and incident 
same_state_tribal <- DR_state_tribal_level %>%
  select(femaDeclarationString, state, incidentId, tribalRequest) %>%
  group_by(incidentId, state) %>%
  mutate(number_in_state = n())

keep <- same_state_tribal %>%
  filter(number_in_state > 1)  # keeping only state-incidents that have both a tribal and non-tribal declaration — these are the direct comparison cases

# filtering for just those id'ed in "keep" 
# this one shows names of storms
DR_state_tribal_level_same_state_Weather <- DR_state_tribal_level %>%
  filter(femaDeclarationString %in% keep$femaDeclarationString) %>%
  # took covid related out - though that looks like a potential story too 
  filter(incidentType != "Biological") 

## remove columns that make pivot difficult 
## pivot to better compare data 
DR_state_tribal_level_same_state_Weather_difference <- DR_state_tribal_level_same_state_Weather %>%
  select(state, incidentId, incidentType, days_to_declaration, tribalRequest) %>%
  pivot_wider(names_from = tribalRequest, values_from = days_to_declaration)  # puts tribal and non-tribal lag side by side in the same row for easy comparison


View(DR_state_tribal_level_same_state_Weather_difference)

library(ggplot2)

ggplot(DR_state_tribal_level_same_state_Weather) + 
  geom_point(aes(x = incidentBeginDate, y = days_to_declaration, group = tribalRequest, col = tribalRequest), alpha = 0.7)
```


```{r write-csv}
write.csv(helene_state_summary, "helene_state_summary.csv", row.names = FALSE)
write.csv(tribal_DR, "tribal_DR.csv", row.names = FALSE)
write.csv(DR_state_tribal_level_same_state_Weather, "DR_state_tribal_level_same_state_Weather.csv", row.names = FALSE)
```

##Overall Trends

```{r overall-trend}
#mapping out overall trend in storm declarations

storms <- DR_state_level|>
  filter(incidentType %in% c("Tropical Storm", "Hurricane"))

# quick exploratory plot before the cleaned-up version below
ggplot(storms) +
  geom_point(aes(x = incidentBeginDate, y = days_to_declaration)) +
  geom_smooth(aes(x = incidentBeginDate, y = days_to_declaration))

write.csv(storms, "overall_trend.csv")

```

```{r plot}
#plotting overall trends for trop. storms and hurricanes

storms <- DR_state_level %>%
  filter(incidentType %in% c("Tropical Storm", "Hurricane"))

ggplot(storms, aes(x = incidentBeginDate, y = days_to_declaration)) +
  geom_point(color = "steelblue", alpha = 0.4, size = 1.5) +
  geom_smooth(color = "black", se = TRUE) +  # se = TRUE shades the confidence interval around the trend line
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "Hurricane & Tropical Storm Declaration Lag Over Time",
    subtitle = "All major disaster declarations (DR); one point per state-level declaration",
    x        = NULL,
    y        = "Days from Incident Begin to Declaration"
  ) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

write.csv(storms, "storms.csv", row.names = FALSE)

```

## Conclusions/Answers

FEMA response for Helene was actually not terrible -- most do not show a delay surpassing ~1 week. I feel like that may be explainable because of what my source told me, which is that for state-level funding/response, there is a lot of paperwork and bureaucracy involved in that which could explain delays. Or the state simply took longer to expend their own resources. KY and WV are big outliers with 60 days between incident start and dec date. Every county listed had the same delay #.

Compared to other storms, Helene response was on the faster end, most similar to Hurricane Matthew in 2016. However, Helene's max declaration date of 75 days in WV is the highest of any storm I selected.

Something interesting too and something I'd like to explore further is why tribal reservations, especially those in FL, experience delays in their declarations in comparison to the rest of the state. This was at its worst during Hurricane Ida, which had a 59 day delay.

Overall, I wouldn't be comfortable making the claim that FEMA has an increasing pattern of a lack of disaster response. Delays seem to be isolated to certain storms, and I suspect there are more political factors involved with Hurricane Helene and FEMA issues behind public outcry against FEMA following the disaster. I'd say my hypothesis that there was a concrete pattern emerging is not all the way true, but it is something that can't be simplified to looking at declaration dates alone because of other moving pieces in disaster response.
