---
title: "FEMA Data Doc"
output: html_notebook
---

## Question: Is there an emerging pattern of increasing delays for federal disaster response?

## Data Sets: Disaster Declarations Summaries v2: <https://drive.google.com/file/d/1YhwtLE-e428HYR-CLruthPc3VHoWb2cg/view?usp=sharing>

## Methodology

I used FEMA's Disaster Declarations Summaries v2, a publicly available dataset spanning the Eisenhower administration to the present, where each row represents a county-level disaster designation. The variable days_to_declaration measures the number of calendar days between incidentBeginDate and declarationDate using difftime() in R, restricted to Major Disaster Declarations only (declarationType == "DR") as these usually indicate more broad federal assistance.

Because each row is a county designation rather than a unique disaster, the data needed to be cleaned. A check of ~3,000 state-incident combinations found only 13 with variation in declaration dates within a single state, which were nearly all attributable to tribal nations receiving separate, typically later, declarations. My analysis groups at the state level and takes the earliest declaration date within each state-incident group — tribal declarations tend to come later and are treated as a separate phenomenon. I keep county-level rows to report minimum, median, and maximum lag per state rather than reducing to a single number. Median is used as the primary statistic over mean to try and avoid a single outlier county disproportionately skewing state-wide data.

Hurricane Helene is isolated by incidentId rather than name string, as a string search for "helene" returns rows from an older unrelated storm since FEMA reuses storm names. I chose Hurricanes Camille, Floyd, Frances, Ivan, Jeanne, Harvey, Ida, Matthew, and Idalia as comparable storms because of their similar rapid intensification and extreme rainfall. For the comparable storms analysis, I tried both a median and a simple unweighted mean to compare — a weighted average by population or damage estimates would be more rigorous but would probably require joining an additional dataset.

For tribal declarations specifically, I filtered using both the tribalRequest flag and a string search across designatedArea for keywords like "indian nation," "indian reservation," "pueblo," and "tribe of." I then separated tribal and non-tribal declarations within the same state-incident to compare lag differences directly, pivoting the data wide to put tribal and non-tribal days side by side. I also isolated Florida specifically given it has a clear compare/contrast case within the same state. One NA that appeared in the unweighted mean chart turned out to be Hurricane Ian's Seminole Tribe of Florida declaration, which shared an incidentId with the broader Ian declaration — I kept it labeled separately as "Ian - Seminole (2022)" in the comparable storms analysis.

I also looked at the overall trend over time for all hurricane and tropical storm DR declarations, plotting days_to_declaration against incidentBeginDate with a smoothed trend line, then broke it down further by decade and by presidential administration using both mean and median.

Two key limitations throughout are that the lag clock starts at incident begin, but FEMA cannot declare until a state governor formally requests it, meaning some delay may reflect state-level decisions rather than federal response time. The other limitation is that data manually entered before the 1998 launch of the National Emergency Management Information System may be subject to a small percentage of human error.

```{r setup}
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(tidyr)
```

```{r load}
df_FEMA_OG <- read.csv('FEMA_disaster_declarations.csv')
```

```{r clean-dates}
# converting all three date columns from character strings to proper Date objects so
# difftime() and any date math works correctly below
df_FEMA_clean <- df_FEMA_OG %>%
  mutate(
    declarationDate   = as_date(declarationDate),
    incidentBeginDate = as_date(incidentBeginDate),
    incidentEndDate   = as_date(incidentEndDate)
  )
```

---

## Checking incidentId vs. string filtering (Professor's note)

Filtering by name string catches an older storm also named Helene. Used `incidentId`.

```{r helene-id-vs-string}
helene_by_id <- df_FEMA_OG %>%
  filter(incidentId == '2024092301')

helene_by_string <- df_FEMA_OG %>%
  filter(grepl("helene", declarationTitle, ignore.case = TRUE))

#since helene by string has more, let's see what it has that's not in the list filtered by ID number

helene_missing_in_id <- helene_by_string %>%
  filter(!id %in% helene_by_id$id)

## and vice versa just in case 

helene_missing_in_string <- helene_by_id %>%
  filter(!id %in% helene_by_string$id)


#filtering using the string includes rows from another storm named helene before!!

cat("Rows caught by string but not ID (older storm):", nrow(helene_missing_in_id), "\n")
cat("Rows caught by ID but not string:", nrow(helene_missing_in_string), "\n")
```

---

## Days to Declaration — All DR Declarations

```{r all-dr}
# restricting to Major Disaster Declarations and creating column that calculates days between incident start and declaration
all_DR_data <- df_FEMA_clean %>%
  filter(declarationType == 'DR') %>%
  mutate(days_to_declaration = as.numeric(difftime(declarationDate, incidentBeginDate, units = "days")))
```

```{r comparable-storms-naming}
# storm names used for string-matching across chunks — defined once here to stay consistent
comparable_names <- c("camille", "floyd", "frances", "ivan", "jeanne",
                      "harvey", "ida", "matthew", "helene")

```

```{r all-dr-comparable}
# quick check to confirm which incidentIds the name-string filter actually returns
# before committing to filtering by ID to catch duplicates or mismatches
DR_data_comparable <- all_DR_data %>%
  filter(incidentType %in% c("Hurricane", "Tropical Storm")) %>%
  filter(str_detect(str_to_lower(declarationTitle),
                    paste(comparable_names, collapse = "|"))) %>%
  distinct(declarationTitle, incidentId, incidentBeginDate) %>%
  arrange(incidentBeginDate)

DR_data_comparable

```

```{r dr-states-incidents}
# check whether declaration dates vary within a state for a given incident
# If count == 1 for all rows, every county in that state-incident got the same date
# only 13 out of ~3,000 state-incident combos have variation

DR_states_incidents <- all_DR_data %>%
  select(femaDeclarationString, state, incidentId, incidentBeginDate, declarationDate) %>%
  unique() %>%
  group_by(state, incidentId) %>%
  summarise(count = n(), .groups = "drop")

# the 13 exceptions
DR_states_incidents %>%
  filter(count > 1) %>%
  arrange(desc(count))
```

```{r check-exceptions-tribal}
# nearly all exceptions appear to be tribal declarations with a
# later date.

exception_ids <- DR_states_incidents %>%
  filter(count > 1) %>%
  pull(incidentId)

# pulling the full rows for each exception to look at designatedArea and tribalRequest
# and confirm that late dates are tribal rather than errors
all_DR_data %>%
  filter(incidentId %in% exception_ids) %>%
  select(femaDeclarationString, state, incidentId, designatedArea,
         incidentBeginDate, declarationDate, days_to_declaration, tribalRequest) %>%
  unique() %>%
  arrange(incidentId, declarationDate)
```

Since within-state date variation is minimal and largely driven by tribal declarations,
I grouped at the state level for the main analysis.

```{r dr-state-level}
# one row per state-incident: take the earliest declaration date within each group
# (tribal declarations tend to be later and are a separate phenomenon)

DR_state_level <- all_DR_data %>%
  select(femaDeclarationString, state, incidentId, incidentType,
         declarationTitle, incidentBeginDate, declarationDate,
         days_to_declaration, tribalRequest, fyDeclared) %>%
  unique() %>%
  group_by(state, incidentId) %>%
  slice_min(declarationDate, n = 1, with_ties = FALSE) %>%  # with_ties = FALSE avoids duplicate rows when two counties share the exact same earliest date
  ungroup()
```

---

## Hurricane Helene

```{r helene-dr}
# isolating Helene by incidentId rather than name string (see helene-id-vs-string chunk above)
helene_data_DR <- df_FEMA_clean %>%
  filter(declarationType == 'DR') %>%
  filter(incidentId == '2024092301') %>%
  mutate(days_to_declaration = as.numeric(difftime(declarationDate, incidentBeginDate, units = "days")))
```

```{r helene-state-summary}
# summarizing at the state level rather than county to get a cleaner picture —
# min/max/median gives a sense of spread within each state without flattening it to one number
helene_state_summary <- helene_data_DR %>%
  group_by(state) %>%
  summarise(
    n_counties = n(),
    min_lag    = min(days_to_declaration),
    median_lag = median(days_to_declaration),
    max_lag    = max(days_to_declaration),
    .groups    = "drop"
  ) %>%
  arrange(median_lag)

helene_state_summary

write.csv(helene_state_summary, "helene_state_summary.csv")

```

```{r helene-plot}
# I heavily used the help of Claude here to help me chart what I found visually. I think for KY and WV being a lot later that might be because of delayed effects? Or maybe that's something to look further into.
# bar height shows median lag in response. States where min and max are far apart
# probably need closer inspection where one county may have had an unusually delayed declaration
helene_state_summary %>%  
ggplot(aes(x = reorder(state, -median_lag), y = median_lag, fill = median_lag)) +
  geom_col() +
  geom_text(aes(label = median_lag), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +  # color encodes severity of delay — longer lags shade toward red
  labs(
    title    = "Hurricane Helene: Median Days to DR Declaration by State",
    subtitle = "Incident began 2024-09-23; bar shows median, see helene_state_summary for min/max",
    x        = NULL,
    y        = "Median Days to Declaration"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Comparable Storms (grouped by incidentId)

Rather than filtering by name string, I grouped all DR declarations by `incidentId`
and then label the storms of interest.

```{r comparable-storms}
#trying to include storms to compare any delays between helene and past hurricanes
comparable_ids <- c(
  "1999091407",                     # Floyd (1999)
  "8091000001",                     # Frances/Georges (1998)
  "2004082901",                     # Frances (2004)
  "2004090502",                     # Ivan (2004)
  "2004091501",                     # Jeanne (2004)
  "2009110901",                     # Ida (2009)
  "2016092802",                     # Matthew (2016)
  "2017082301",                     # Harvey (2017)
  "2021082601",                     # Ida (2021)
  "2022092201",                     # Ian (2022)
  "2023082403",                     # Idalia (2023)
  "2024092301"                      # Helene (2024)
)

comparable_storms <- all_DR_data %>%
  filter(incidentId %in% comparable_ids) %>%
  mutate(
    storm_label = case_when(
  incidentId == "1999091407" ~ "Hurricane Floyd (1999)",
  incidentId == "8091000001" ~ "Tropical Storm Frances/Hurricane Georges (1998)",
  incidentId == "2004082901" ~ "Hurricane Frances (2004)",
  incidentId == "2004090502" ~ "Hurricane Ivan (2004)",
  incidentId == "2004091501" ~ "Hurricane Jeanne (2004)",
  incidentId == "2009110901" ~ "Hurricane Ida (2009)",
  incidentId == "2016092802" ~ "Hurricane Matthew (2016)",
  incidentId == "2017082301" ~ "Hurricane Harvey (2017)",
  incidentId == "2021082601" ~ "Hurricane Ida (2021)",
  #Here I realized two different "Ians" were under the same incident ID. I separated those here by designated area to separate tribal declarations, since they are considered separate from the state by FEMA.
  incidentId == "2022092201" & str_detect(str_to_lower(designatedArea), "seminole") ~ "Hurricane Ian - Seminole (2022)",
  incidentId == "2022092201" ~ "Hurricane Ian (2022)",
  incidentId == "2023082403" ~ "Hurricane Idalia (2023)",
  incidentId == "2024100501" ~ "Hurricane Milton (2024)",
  incidentId == "2024092301" ~ "Hurricane Helene (2024)"
)
  ) %>%
  distinct(incidentId, designatedArea, state, .keep_all = TRUE) %>%  # deduplicating at the county level before summarizing so a county appearing in multiple rows doesn't inflate the count
  group_by(storm_label) %>%
  summarise(
  #to count number of counties & filter for unique instances with distinct to prevent duplicates
    n_counties = n(),
    min_lag    = min(days_to_declaration),
    median_lag = median(days_to_declaration),
    max_lag    = max(days_to_declaration),
    .groups    = "drop"
  )
```

```{r comparable-plot}
#plotting those comparable storms
comparable_storms %>%
  mutate(year = as.numeric(str_extract(storm_label, "\\d{4}"))) %>%  
  ggplot(aes(x = reorder(storm_label, year), y = median_lag)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(median_lag, 0)), hjust = -0.2, size = 3.5) +
  coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + 
  labs(
    title    = "Median Declaration Lag for Storms Comparable to Helene",
    subtitle = "DR declarations only; median across all designated counties",
    x        = NULL,
    y        = "Median Days from Incident Begin to Declaration"
  ) +
  theme_minimal()
```

```{r comparable-unweighted}
#i wanted to include something that didn't take into account the number of counties in a state just in case it was useful to compare
comparable_storms_unweighted <- all_DR_data %>%
  filter(incidentId %in% comparable_ids) %>%
  filter(days_to_declaration >= 0) %>%
  mutate(
   storm_label = case_when(
  incidentId == "1999091407" ~ "Hurricane Floyd (1999)",
  incidentId == "8091000001" ~ "Tropical Storm Frances/Hurricane Georges (1998)",
  incidentId == "2004082901" ~ "Hurricane Frances (2004)",
  incidentId == "2004090502" ~ "Hurricane Ivan (2004)",
  incidentId == "2004091501" ~ "Hurricane Jeanne (2004)",
  incidentId == "2009110901" ~ "Hurricane Ida (2009)",
  incidentId == "2016092802" ~ "Hurricane Matthew (2016)",
  incidentId == "2017082301" ~ "Hurricane Harvey (2017)",
  incidentId == "2021082601" ~ "Hurricane Ida (2021)",
  incidentId == "2022092201" & str_detect(str_to_lower(designatedArea), "seminole") ~ "Hurricane Ian - Seminole (2022)",
  incidentId == "2022092201" ~ "Hurricane Ian (2022)",
  incidentId == "2023082403" ~ "Hurricane Idalia (2023)",
  incidentId == "2024100501" ~ "Hurricane Milton (2024)",
  incidentId == "2024092301" ~ "Hurricane Helene (2024)"
)
  ) %>%
  filter(!is.na(storm_label)) %>%  # drops Ian's Seminole rows, which don't match any label — they show up as NA and cause the chart gap identified in the na chunk below
  group_by(storm_label) %>%
  summarise(
    n_counties = n(),
    mean_lag   = round(mean(days_to_declaration), 1),
    .groups    = "drop"
  )
```

```{r comparable-plot-unweighted}
comparable_storms_unweighted %>%
  mutate(year = as.numeric(str_extract(storm_label, "\\d{4}"))) %>%
  ggplot(aes(x = reorder(storm_label, year), y = mean_lag)) +
  geom_col(fill = "firebrick", alpha = 0.8) +
  geom_text(aes(label = mean_lag), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Unweighted Mean Declaration Lag for Storms Comparable to Helene",
    subtitle = "DR declarations only; simple mean across all designated counties — compare with median chart above",
    x        = NULL,
    y        = "Mean Days from Incident Begin to Declaration"
  ) +
  theme_minimal()
```
#Having trouble figuring out why one is showing up as NA. I used code below to try and filter for what was causing the problem and I guess Ian is Hurricane Ian + Hurricane Ian Seminole Tribe of Florida, both under the same incidentID.

```{r na}
# isolating whichever rows didn't match any storm_label in the unweighted chunk —
# anything that falls through all the case_when conditions comes out as NA
df_findna <- all_DR_data %>%
  filter(incidentId %in% comparable_ids) %>%
  filter(days_to_declaration >= 0) %>%
  mutate(
    storm_label = case_when(
      incidentId %in% c("69034", "69035", "69043") ~ "Camille (1969)",
      incidentId == "1999091407" ~ "Floyd (1999)",
      incidentId == "8091000001" ~ "Frances/Georges (1998)",
      incidentId == "2004082901" ~ "Frances (2004)",
      incidentId == "2004090502" ~ "Ivan (2004)",
      incidentId == "2004091501" ~ "Jeanne (2004)",
      incidentId == "2009110901" ~ "Ida (2009)",
      incidentId == "2016092802" ~ "Matthew (2016)",
      incidentId == "2017082301" ~ "Harvey (2017)",
      incidentId == "2021082601" ~ "Ida (2021)",
      incidentId == "2023082403" ~ "Idalia (2023)",
      incidentId == "2024092301" ~ "Helene (2024)"
      # Ian (2022) intentionally omitted here to surface the NA
    )
  ) %>%
  filter(is.na(storm_label)) %>%
  distinct(incidentId, declarationTitle)

```

```{r tribal}
# double-filtering for tribal designations: first by tribalRequest,
# then by string-matching designatedArea for common tribal naming conventions
tribal_DR <- all_DR_data %>%
  filter(
  tribalRequest == 1 |
  str_detect(str_to_lower(designatedArea), "indian nation|indian reservation|indian tribe|band of cherokee|band of indian|pueblo|tribe of")
)%>%
  filter(
    str_detect(str_to_lower(declarationTitle),
               paste(comparable_names, collapse = "|")) | incidentId == "2024092301"  # incidentId added as fallback since "helene" is in comparable_names but this makes the Helene filter more specific
  ) %>%
  group_by(state, designatedArea, declarationTitle, fyDeclared) %>%
  summarise(
    n_declarations = n(),
    min_lag        = min(days_to_declaration),
    median_lag     = median(days_to_declaration),
    max_lag        = max(days_to_declaration),
    .groups        = "drop"
  ) %>%
  arrange(desc(median_lag))

tribal_DR
```


```{r tribal-plot}
tribal_DR %>%
  mutate(area_label = paste0(designatedArea, " (", declarationTitle, ")")) %>%  # combining area and storm name into one label so each bar is fully identifiable without a legend
  ggplot(aes(x = reorder(area_label, median_lag), y = median_lag, fill = median_lag)) +
  geom_col() +
  geom_text(aes(label = median_lag), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Declaration Lag for Tribal Areas",
    subtitle = "Helene + comparable storms",
    x        = NULL,
    y        = "Median Days to Declaration"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 7))
```

```{r}
# adding in tribal to the group_by
DR_state_tribal_level <- all_DR_data %>%
  select(femaDeclarationString, state, incidentId, incidentType,
         declarationTitle, incidentBeginDate, declarationDate,
         days_to_declaration, tribalRequest, fyDeclared) %>%
  mutate(tribalRequest = ifelse(tribalRequest == 1, 'Days - Tribal', 'Days - Not Tribal')) %>%  # converting binary flag to readable labels for use as pivot column names below
  unique() %>%
  group_by(state, incidentId, tribalRequest) %>%
  slice_min(declarationDate, n = 1, with_ties = FALSE) %>%  # taking earliest date within each state-incident-tribalRequest group, same logic as DR_state_level
  ungroup()

# finding when more than one declaration for a given state and incident 
same_state_tribal <- DR_state_tribal_level %>%
  select(femaDeclarationString, state, incidentId, tribalRequest) %>%
  group_by(incidentId, state) %>%
  mutate(number_in_state = n())

keep <- same_state_tribal %>%
  filter(number_in_state > 1)  # keeping only state-incidents that have both a tribal and non-tribal declaration — these are the direct comparison cases

# filtering for just those id'ed in "keep" 
# this one shows names of storms
DR_state_tribal_level_same_state_Weather <- DR_state_tribal_level %>%
  filter(femaDeclarationString %in% keep$femaDeclarationString) %>%
  # took covid related out - though that looks like a potential story too 
  filter(incidentType != "Biological") 

## remove columns that make pivot difficult 
## pivot to better compare data 
DR_state_tribal_level_same_state_Weather_difference <- DR_state_tribal_level_same_state_Weather %>%
  select(state, incidentId, incidentType, days_to_declaration, tribalRequest) %>%
  pivot_wider(names_from = tribalRequest, values_from = days_to_declaration)  # puts tribal and non-tribal lag side by side in the same row for easy comparison


View(DR_state_tribal_level_same_state_Weather_difference)

library(ggplot2)

ggplot(DR_state_tribal_level_same_state_Weather) + 
  geom_point(aes(x = incidentBeginDate, y = days_to_declaration, group = tribalRequest, col = tribalRequest), alpha = 0.7)
```

```{r}


```


```{r write-csv}
write.csv(helene_state_summary, "helene_state_summary.csv", row.names = FALSE)
write.csv(comparable_storms, "comparable_storms.csv", row.names = FALSE)
write.csv(comparable_storms_unweighted, "comparable_storms_unweighted.csv", row.names = FALSE)
write.csv(tribal_DR, "tribal_DR.csv", row.names = FALSE)
write.csv(DR_state_tribal_level_same_state_Weather, "DR_state_tribal_level_same_state_Weather.csvsevel_same_state_Weather.csv", row.names = FALSE)
```

```{r overall-trend}
#overall trend df + plot from analysis feedback

storms <- DR_state_level|>
  filter(incidentType %in% c("Tropical Storm", "Hurricane"))

# quick exploratory plot before the cleaned-up version below
ggplot(storms) +
  geom_point(aes(x = incidentBeginDate, y = days_to_declaration)) +
  geom_smooth(aes(x = incidentBeginDate, y = days_to_declaration))

write.csv(storms, "overall_trend.csv")

```

```{r plot}
#plotting overall trends for trop. storms and hurricanes

storms <- DR_state_level %>%
  filter(incidentType %in% c("Tropical Storm", "Hurricane"))

ggplot(storms, aes(x = incidentBeginDate, y = days_to_declaration)) +
  geom_point(color = "steelblue", alpha = 0.4, size = 1.5) +
  geom_smooth(color = "black", se = TRUE) +  # se = TRUE shades the confidence interval around the trend line
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title    = "Hurricane & Tropical Storm Declaration Lag Over Time",
    subtitle = "All major disaster declarations (DR); one point per state-level declaration",
    x        = NULL,
    y        = "Days from Incident Begin to Declaration"
  ) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

write.csv(storms, "storms.csv", row.names = FALSE)

```

```{r storms-decades}
#Wanted to get a sense of what this looks like breaking down by decade and then by presidential administration

# decade summary
decade_summary <- storms %>%
  mutate(
    decade = paste0(floor(year(incidentBeginDate) / 10) * 10, "s")  # floor division rounds down to the decade start, then "s" is appended for the label (e.g., 1990 -> "1990s")
  ) %>%
  group_by(decade) %>%
  summarise(
    avg_days = round(mean(days_to_declaration, na.rm = TRUE), 1),
    median_days = round(median(days_to_declaration, na.rm = TRUE), 1),
    n_declarations = n()
  ) %>%
  arrange(decade)

# Administration summary
admin_order <- c(
  "Eisenhower (R)", "Kennedy (D)", "Johnson (D)", "Nixon (R)",
  "Ford (R)", "Carter (D)", "Reagan (R)", "H.W. Bush (R)",
  "Clinton (D)", "W. Bush (R)", "Obama (D)", "Trump I (R)",
  "Biden (D)", "Trump II (R)"
)

admin_summary <- storms %>%
  mutate(
    administration = case_when(
      year(incidentBeginDate) >= 1953 & year(incidentBeginDate) <= 1961 ~ "Eisenhower (R)",
      year(incidentBeginDate) >= 1961 & year(incidentBeginDate) <= 1963 ~ "Kennedy (D)",
      year(incidentBeginDate) >= 1963 & year(incidentBeginDate) <= 1969 ~ "Johnson (D)",
      year(incidentBeginDate) >= 1969 & year(incidentBeginDate) <= 1974 ~ "Nixon (R)",
      year(incidentBeginDate) >= 1974 & year(incidentBeginDate) <= 1977 ~ "Ford (R)",
      year(incidentBeginDate) >= 1977 & year(incidentBeginDate) <= 1981 ~ "Carter (D)",
      year(incidentBeginDate) >= 1981 & year(incidentBeginDate) <= 1989 ~ "Reagan (R)",
      year(incidentBeginDate) >= 1989 & year(incidentBeginDate) <= 1993 ~ "H.W. Bush (R)",
      year(incidentBeginDate) >= 1993 & year(incidentBeginDate) <= 2001 ~ "Clinton (D)",
      year(incidentBeginDate) >= 2001 & year(incidentBeginDate) <= 2009 ~ "W. Bush (R)",
      year(incidentBeginDate) >= 2009 & year(incidentBeginDate) <= 2017 ~ "Obama (D)",
      year(incidentBeginDate) >= 2017 & year(incidentBeginDate) <= 2021 ~ "Trump I (R)",
      year(incidentBeginDate) >= 2021 & year(incidentBeginDate) <= 2025 ~ "Biden (D)",
      year(incidentBeginDate) >= 2025 ~ "Trump II (R)",
      TRUE ~ "Unknown"
    ),
    administration = factor(administration, levels = admin_order)  # setting factor levels to admin_order forces the x-axis into chronological order rather than alphabetical
  ) %>%
  group_by(administration) %>%
  summarise(
    avg_days = round(mean(days_to_declaration, na.rm = TRUE), 1),
    median_days = round(median(days_to_declaration, na.rm = TRUE), 1),
    n_declarations = n()
  )

# Decade chart
ggplot(decade_summary, aes(x = decade)) +
  geom_col(aes(y = avg_days), fill = "steelblue", alpha = 0.7, width = 0.4,
           position = position_nudge(x = -0.22)) +  # nudging mean and median bars in opposite directions so they sit side by side rather than overlapping
  geom_col(aes(y = median_days), fill = "firebrick", alpha = 0.7, width = 0.4,
           position = position_nudge(x = 0.22)) +
  geom_text(aes(y = avg_days, label = avg_days),
            position = position_nudge(x = -0.22), vjust = -0.4, size = 3) +
  geom_text(aes(y = median_days, label = median_days),
            position = position_nudge(x = 0.22), vjust = -0.4, size = 3) +
  labs(
    title = "Hurricane & Tropical Storm Declaration Lag by Decade",
    subtitle = "Blue = mean, Red = median | Major Disaster Declarations only",
    x = NULL,
    y = "Days to Declaration"
  ) +
  theme_minimal()

# Administration chart
ggplot(admin_summary, aes(x = administration)) +
  geom_col(aes(y = avg_days), fill = "steelblue", alpha = 0.7, width = 0.4,
           position = position_nudge(x = -0.22)) +
  geom_col(aes(y = median_days), fill = "firebrick", alpha = 0.7, width = 0.4,
           position = position_nudge(x = 0.22)) +
  geom_text(aes(y = avg_days, label = avg_days),
            position = position_nudge(x = -0.22), vjust = -0.4, size = 3) +
  geom_text(aes(y = median_days, label = median_days),
            position = position_nudge(x = 0.22), vjust = -0.4, size = 3) +
  labs(
    title = "Hurricane & Tropical Storm Declaration Lag by Administration",
    subtitle = "Blue = mean, Red = median | Major Disaster Declarations only",
    x = NULL,
    y = "Days to Declaration"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Conclusions/Answers

FEMA response for Helene was actually not terrible -- most do not show a delay surpassing ~1 week. I feel like that may be explainable because of what my source told me, which is that for state-level funding/response, there is a lot of paperwork and bureaucracy involved in that which could explain delays. Or the state simply took longer to expend their own resources. KY and WV are big outliers with 60 days between incident start and dec date. Every county listed had the same delay #.

Compared to other storms, Helene response was on the faster end, most similar to Hurricane Matthew in 2016. However, Helene's max declaration date of 75 days in WV is the highest of any storm I selected.

Something interesting too and something I'd like to explore further is why tribal reservations, especially those in FL, experience delays in their declarations in comparison to the rest of the state. This was at its worst during Hurricane Ida, which had a 59 day delay.

Overall, I wouldn't be comfortable making the claim that FEMA has an increasing pattern of a lack of disaster response. Delays seem to be isolated to certain storms, and I suspect there are more political factors involved with Hurricane Helene and FEMA issues behind public outcry against FEMA following the disaster. I'd say my hypothesis that there was a concrete pattern emerging is not all the way true, but it is something that can't be simplified to looking at declaration dates alone because of other moving pieces in disaster response.
